A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-15 podc_paper/introduction.tex
related work; made bib entries consistent

A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues,
starting with the classic MS-queue of Michael and Scott \cite{DBLP:conf/podc/MichaelS96}.
Their design uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{list-some}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} dubbed this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{list-some}.

>>>>

A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues.
As well as being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
Their design uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{list-some}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} dubbed this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{list-some}.

### 2022-12-15 podc_paper/related.tex
related work; made bib entries consistent

Michael and Scott \cite{DBLP:conf/podc/MichaelS96} describe the early history of concurrent queue algorithms
in the paper that described their MS-queue \cite{DBLP:conf/podc/MichaelS96}, which
is a lock-free queue that has stood the test of time.
A version of it is
included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
However, as mentioned above, it suffers from the CAS retry problem
which means it is lock-free but not wait-free and has an amortized step complexity
of $\Theta(p)$ per operation.

A number of papers have described ways to cope with the high contention at the front
and back of the MS-queue.
Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} 
added an elimination array that allows an enqueue to pass its value directly
to a concurrent dequeue when the queue is empty or when concurrent operations 
can be linearized to empty the queue.
However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
is still present.
The baskets queue of
Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
is another approach to avoid contention:  concurrent enqueues are grouped into baskets
and operations within a basket can be ordered arbitrarily.
However, if there are $p$ concurrent enqueues in the same basket
the CAS retry problem occurs when they must be ordered into a list using CAS instructions.
Both modifications thus still have $\Omega(p)$ amortized step complexity.

@@

Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free
queue supporting futures. In their queue, operations  return future
objects instead of responses. Later when the response is needed, it
can be evaluated from the future object. They also define a weaker
linearizability condition such that each operation can be linearized
between its invocation and when its future is evaluated. Their idea of
batching allows a sequence of operations to be submitted as a batch
for later execution on the MS-queue. They use some properties of the
queue size before and after a batch, similar to a part of our
work. Their queue is not wait-free: in fact, if the batch sizes are 1,
then the queue is like MS-queue.

>>>>

Michael and Scott \cite{MS98} describe the early history of concurrent queue algorithms
in the paper that described their MS-queue, which
is a lock-free queue that has stood the test of time.
A version of it is
included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
However, as mentioned above, it suffers from the CAS retry problem
which means it is lock-free but not wait-free and has an amortized step complexity
of $\Theta(p)$ per operation.

A number of papers have described ways to reduce contention at the front
and back of the MS-queue.
Moir et al.~\cite{DBLP:conf/spaa/MoirNSS05} 
added an elimination array that allows an enqueue to pass its enqueued value directly
to a concurrent dequeue when the queue is empty or when concurrent operations 
can be linearized to empty the queue.
However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
is still present.
The baskets queue of
Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
is another approach to avoid contention:  concurrent enqueues are grouped into baskets.
If an enqueue fails its CAS, it is put in the basket with the operation that succeeded,
and operations within a basket can be ordered arbitrarily.
However, if there are $p$ concurrent enqueues in the same basket
the CAS retry problem occurs when they order themselves into a list using CAS instructions.
Both modifications thus still have $\Omega(p)$ amortized step complexity.

- MKLL18 batches several of a process's operations together and apply them as a group to the queue but this assumes processes can wait to do a batch all at once.  In worst case batches of size 1 reduces to MS98
- RC17 sketched how to add a helping mechanism to ensure wait-freedom (increases amortized time to Omega($P^2$) for enqueue)
- Kogan Petrank PPoPP 2011 add helping to MS98 queue to make it wait-free; only makes complexity worse

LS08 modified MS98; switched to doubly-linked list to reduce number of successful CASes needed for an op from 2 to 1 in optimistic approach, but could get inconsistent pointers that can be fixed when needed.  Fixing the list, though rare in practice, is expensive, so Omega(nP) amortized time per op.

@@

Milman-Sela et al.~\cite{MKLLP22} designed a lock-free
queue supporting futures. In their queue, operations  return future
objects instead of responses. Later when the response is needed, it
can be evaluated from the future object. They also define a weaker
linearizability condition such that each operation can be linearized
between its invocation and when its future is evaluated. Their idea of
batching allows a sequence of operations to be submitted as a batch
for later execution on the MS-queue. They use some properties of the
queue size before and after a batch, similar to a part of our
work. Their queue is not wait-free: in fact, if the batch sizes are 1,
then the queue is like MS-queue.
